CSE530 Project Final Report\\ Bandwidth-Aware Memory Hierarchy Design with Hybrid Memory Technologies

### 2010-11-18 trunk/paper.tex
version 1.0

\documentclass[10pt,letterpaper,singlecolumn]{article}
\usepackage{setspace}
\doublespacing
\usepackage{latex8}
\usepackage{times}
\usepackage{cite}
\usepackage{subfigure}
\usepackage{psfig}
\usepackage[pdftex]{graphicx}
\usepackage{epsfig}
\usepackage{pstricks}
\usepackage{fullpage}
\usepackage{amsthm}
\usepackage{graphics}
\usepackage{comment}
\usepackage[small,compact]{titlesec}
\usepackage[small,it]{caption}
\usepackage{wrapfig}

@@

Many of modern chip-multiprocessors are designed to perform well on various
applications to achieve high performance by exploiting their inherent
parallelism. Such systems support large number of threads and single instruction
multiple data~(SIMD) execution, which puts a lot of pressure on the memory
system. Memory latency is typically not a bottleneck, since the latency can be
hidden via multi-threading or hardware prefetching. To this end, bandwidth becomes a
potential bottleneck. A high rate of computing often brings in a high rate of
data transitions. In some cases, the working set of an application fits in the
on-die caches, which can typically provide sufficient bandwidth to keep up with
the processing cores. However, if the working set does not fit in the on-die
caches, the main memory needs to provide much of the data. Since the bandwidth
of off-chip main memory is quite limited, applications with such working sets
are potentially bandwidth-bound. Therefore, it is crucial to design the memory
hierarchy to overcome the bandwidth limitation.

It is known that performance improvement of a computing system can be achieved
via multiple memory levels. Adding an extra level of memory, however, can also
help alleviate the bandwidth bottleneck of off-chip memory. Specifically, we
would like to examine (1) the number of levels in the optimal memory hierarchy,
(2) the appropriate memory technology of each level, and (3) the capacity and
bandwidth of each level. We will also explore the energy-efficiency of the
memory hierarchy design constrain, and the memory hierarchy design within a
fixed power budge.\vspace{0.15in}

@@

\begin{figure*}[htbp]
% The "!" means to maintain the aspect ratio.
\centering
\includegraphics[width=3in]{figures/read-bw}
\includegraphics[width=3in]{figures/write-bw}\\

\vspace*{-1in}
\hspace{0.02in}
\makebox[3in][l]{\textcolor [rgb]{0,0,0} {\bf (a)}}
\makebox[3in][l]{\textcolor [rgb]{0,0,0} {\bf (b)}}

\caption{Read and write bandwidths provided by different memory technologies.
(a) Read bandwidth provided by different memory technologes. (b) Write bandwidth
provided by different memory technologies.}
\label{fig:memory-bw}
\end{figure*}

\begin{figure*}[htbp]
% The "!" means to maintain the aspect ratio.
\centering
\includegraphics[width=2in]{figures/read-energy}
\includegraphics[width=2in]{figures/write-energy}
\includegraphics[width=2in]{figures/total-energy}\\

\vspace*{-0.5in}
\hspace{0.02in}
\makebox[2in][l]{\textcolor [rgb]{0,0,0} {\bf (a)}}
\makebox[2in][l]{\textcolor [rgb]{0,0,0} {\bf (b)}}
\makebox[2in][l]{\textcolor [rgb]{0,0,0} {\bf (c)}}

\caption{Dynamic energy consumption with the provided bandwidths of different
  memory technologies. (a) Read energy of different memory technologes. (b)
  Write energy of different memory technologies. (c) Total energy consumption
  (with 10\% write) of different memory technologies.}
\label{fig:memory-energy}
\end{figure*}

First of all, we estimate the read and write bandwidths that can be provided by
different memory technologies. Figure~\ref{fig:memory-bw} shows the results,
with both x- and y-values in \emph{log} scale. The figure illustrates both the
provided read and write bandwidth as a function of memory capacity. Each of the
memory technologies actually provide nearly the same read bandwidths, as is
shown in figure~\ref{fig:memory-bw}(a). On the other hand, a straight forward
observation from figure~\ref{fig:memory-bw}(b) is that the write bandwidth
varies among different memory technologies. The shape of the SRAM write
bandwidth curve is very similar to the read bandwidth curve. The write bandwidth
curves of the other three memory technologies appear to be very different. The
reason is that write latencies of the three non-volatile memories are much
higher than read latencies. Another observation from
figure~\ref{fig:memory-bw}(b) is that the curves cross to each other at
different locations. Based on these two observations, we would like to design a
bandwidth-aware hybrid memory hierarchy, which always provides the high memory
bandwidth with the given capacity. For example, we probably would like to use
1) SRAM when the memory capacity is under 256MB, 2) MRAM or RRAM between
256MB and 1GB (if only bandwidth is considered), and 3) PCRAM when the memory
cacity is over 1GB.

@@

>>>>

\documentclass[10pt,letterpaper,singlecolumn]{article}
\usepackage{setspace}
\doublespacing
\usepackage{latex8}
\usepackage{times}
\usepackage{cite}
\usepackage{subfigure}
\usepackage{psfig}
\usepackage[pdftex]{graphicx}
\usepackage{epsfig}
\usepackage{pstricks}
\usepackage{fullpage}
\usepackage{amsthm}
\usepackage{graphics}
\usepackage{comment}
\usepackage[small,compact]{titlesec}
\usepackage[small,it]{caption}
\usepackage{wrapfig}
\usepackage{enumerate}
\usepackage{multirow}

@@

Many of modern chip-multiprocessors are designed to perform well on various applications to achieve high performance by exploiting their inherent parallelism. Such systems support large number of threads and single instruction multiple data~(SIMD) execution, which puts a lot of pressure on the memory system. Typically, memory latency is not a bottleneck, since the latency can be hidden via multi-threading or hardware prefetching. To this end, bandwidth becomes a potential bottleneck. A high rate of computing often brings in a high rate of data transitions. In some cases, the working set of an application fits in the on-die caches, which can typically provide sufficient bandwidth to keep up with the processing cores. However, if the working set does not fit in the on-die caches, the main memory needs to provide much of the data. Since the bandwidth of off-chip main memory is quite limited, applications with such working sets are potentially bandwidth-bound. Therefore, it is crucial to design the memory hierarchy to overcome the bandwidth limitation.

It is known that performance improvement of a computing system can be achieved via multiple memory levels. Adding an extra level of memory, however, can also help alleviate the bandwidth bottleneck of off-chip memory. Specifically, we examine (1) the number of levels in the optimal memory hierarchy, (2) the appropriate memory technology of each level, and (3) the capacity and bandwidth of each level. We will also explore the energy-efficiency of the memory hierarchy design constrain, and the memory hierarchy design within a fixed power budge.

Emerging memory that provides fast random access, high storage density, and non-volatility within one memory technology becomes possible  due to the emergence of  various new non-volatile memory~(NVM) technologies, such as spin-torque-transfer memory~(STT-RAM), phase-change memory~(PCRAM), and resistive memory~(RRAM).  Our goal in this project to to leverage these NVM in devising a memory system with multiple layers of the existing memory hierarchy for modern computers, to supply a wide spectrum of final products that cover from a highly latency-optimized microprocessor cache to a highly density-optimized secondary storage device.

The report is organized as follows. First, we present the motivation of design a memory system with better memory bandwidth. Second, we make a quick summary of current emerging memory technologies. Then we use our NVsim, a circuit-level NVM model to evaluate the read latency, write latency and dynamic energy of different memory technologies. After that, we perform the cycle-accurate architectural simulation to find the best memory system design for different applications. Finally we conclude our work and propose potential future work.

@@

\begin{figure*}[htbp]
% The "!" means to maintain the aspect ratio.
\centering
\includegraphics[width=6in]{figures/RAM_bandwidth}
\caption{Read and write bandwidths provided by different memory technologies.
(a) Read bandwidth provided by different memory technologes. (b) Write bandwidth
provided by different memory technologies.}
\label{fig:memory-bw}
\end{figure*}

First of all, we estimate the read and write bandwidths that can be provided by different memory technologies. Figure~\ref{fig:memory-bw} shows the results, with both x- and y-values in \emph{log} scale. The figure illustrates both the
provided read and write bandwidth as a function of memory capacity. Each of the memory technologies actually provide nearly the same read bandwidths, as is shown in figure~\ref{fig:memory-bw}(a). On the other hand, a straight forward observation from figure~\ref{fig:memory-bw}(b) is that the write bandwidth varies among different memory technologies. The shape of the SRAM write bandwidth curve is very similar to the read bandwidth curve. The write bandwidth curves of the other three memory technologies appear to be very different. The reason is that write latencies of the three non-volatile memories are much higher than read latencies. Another observation from figure~\ref{fig:memory-bw}(b) is that the curves cross to each other at different locations.

\begin{figure*}[htbp]
% The "!" means to maintain the aspect ratio.
\centering
\includegraphics[width=6in]{figures/RAM_latency}
\caption{Latency of different memory technologies. (a) Read Latency. (b) Write Latency.}
\label{fig:memory-latency}
\includegraphics[width=6in]{figures/RAM_energy}
\caption{Dynamic energy consumption with the provided bandwidths of different
  memory technologies. (a) Dynamic Energy with 20\% write.  (b) Dynamic Energy with 50\% write. }
\label{fig:memory-energy}
\end{figure*}

Being aware that the read and write latencies of NVM are asymmetric, we consider the read latency at first. In NVsim, we divide the entire cache read latency into the following components:
\begin{itemize}
\item (a)	H-tree input delay
\item (b)	Decoder + word-line delay
\item (c)	Bit-line delay + Sense Amplifier delay
\item (d)	Comparator Delay (for tag part only)
\item (e)	H-tree output delay
\end{itemize}
H-tree latency (a) and (f) are mainly determined by the RC delay of global wires, which is positive proportional to the area of memory macro. Sensing delay (d) is related to the read noise margin of memory cell that is affected by off/on resistance ratio. Figure~\ref{fig:memory-latency} (a) illustrates the read latency of different memories. We can see that sensing delay dominates the read latency of NVM at small capacity so that PCRAM (with the largest resistance window) is faster than RRAM and MRAM (with the smallest resistance window). While H-tree delay unveils at large capacity so that RRAM (with the smallest cell size) becomes faster than PRAM and MRAM (with the biggest cell size). The read latency of SRAM bank will increase rapidly after 128MB due to large area.

Write latency of NVM are almost dominated by the write pulse width. In this work we assume 10ns, 20ns, 100ns for MRAM, RRAM and PCRAM. While write latency of SRAM and eDRAM is a function of capacity, as similar to the read latency. The results in ~\ref{fig:memory-latency} (b) indicates that NVM is suitable for memory with large capacity.

Figure~\ref{fig:memory-energy} has demonstrated the dynamic energy of different memory technologies when 20\% and 50\% write access are assumed. eDRAM will be better than SRAM in terms of dynamic energy after 16MB and this is verified by IBM Power7 L3 cache. The cross-point between NVM and SRAM/eDRAM is postponed for 50\% write than 20\% write.

Based on these observations, we would like to design a bandwidth-aware hybrid memory hierarchy, which always provides the high memory bandwidth with the given capacity. For example,
\begin{itemize}
\item eDRAM has better latency and enegry than SRAM when capacity is larger than 16MB.
\item MRAM is more competitive to SRAM and eDRAM when capacity is larger than 128MB.
\item PCRAM has serious endurance issue and is targeted as main memory replacement.
\item RRAM might fit into cache hierarchy as last level cache replacement when there multiple levels of cache for applications with very few writes.
\end{itemize}

@@

We use Simics~\cite{Magnusson:2002:simics-orig-paper} as the simulator in our
experiments. It is configured to model an eight-core CMP. Each core is in-order, and
is similar to UltraSPARC III architecture. The frequency of each core is set to
be 1GHz. Table~\ref{table:cmp-config} lists the detailed parameters of the
baseline. Our baseline contains 8 private L1 instruction and data caches
respectively. Since we only evaluate shared cache hierarchies, each of L1 cache
is fixed to be SRAM-based, and of 64KB capacity. As regard to lower level
caches, we evaluate two cases, pure SRAM-based and hybrid caches with various
memory technologies, including SRAM, MRAM, RRAM, and eDRAM. By evaluating both
cases, we would like to find out optimal cache design that leads to the peak
performance, i.e., the number of cache levels that is reqired, memory technology
used for each level, and capacity of each level.

\begin{table}[t]
  \centering
  \caption{Baseline CMP configuration.}
  \vspace{0.1in}
  \begin{tabular}{l|c}
    \hline
    \multicolumn{2}{l}{\textbf{Core}}\\
    \hline
    No. of cores & 8\\
    \hline
    Frequency & 1GHz\\
    \hline
    Core architecture & in-order, 14-stage pipline, 2 ALUs, 2 FPUs\\
    \hline
    \multicolumn{2}{l}{\textbf{Memory}}\\
    \hline
    \multirow{2}{*}{Private caches} & L1 I-cache: SRAM, 8 x 64KB, 64B line, 2-way, write-through \\
                                    & L1 D-cache: SRAM, 8 x 64KB, 64B line, 2-way, write-through \\
    \hline
    \multirow{2}{*}{Shared caches} & Case 1: Pure SRAM, 64B line, 8-way, write-back \\
                                   & Case 2: Hybrid~(SRAM, MRAM, RRAM, eDRAM), 64B line, 8-way, write-back\\
    \hline
    Main memory & 4GB\\
    \hline
  \end{tabular}
  \label{table:cmp-config}
\end{table}

The benchmarks are selected from PARSEC benchmark suite~\cite{PARSEC:2008} with
multithreaded programs, which focus on emerging workloads that are designed to
be representative of next-generation shared-memory programs for CMPs. Since the
performance of different memory technologies are closely related to read and
write intensities, we selected some workloads that vary in the average numbers
of L2 cache read per thousand instructions~(RPKI) and write per thousand
instructions~(WPKI), which are listed in
Table~\ref{table:benchmarks}.\vspace{0.3in}

\begin{table}[t]
  \centering
  \caption{Characteristics of selected benchmarks.}
  \vspace{0.1in}
  \begin{tabular}{l|c|c||l|c|c}
    \hline
    \textbf{Benchmarks} & \textbf{RPKI} & \textbf{WPKI} &
    \textbf{Benchmarks} & \textbf{RPKI} & \textbf{WPKI}\\
    \hline
    blackscholes & 22.8 & 61.7 & ferret & 5.7 & 173.4\\
    \hline
    bodytrack & 5.4 & 139.5 & fluidanimate & 2.6 & 70.1\\
    \hline
    canneal & 5.4 & 29.3 & streamcluster & 17.3 & 16.9\\
    \hline
    facesim & 6.0 & 102.4 & swaptions & 2.6 & 121.4\\
    \hline
  \end{tabular}
  \label{table:benchmarks}
\vspace{0.5in}
\end{table}
